---
title: "3_Peak Comparison"
author: "Laurin Schaller"
date: "2023-06-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(rtracklayer)
  library(GenomicRanges)
})
```


## Subsetting to chromosome 1

First we want to subset all the bigBed-files we downloaded from ENCODE to chromosome 1 for reduced sizes.
The following function uses a pythonic approach to looping all the histone marks which have  the same folder names. Within these folders it gets all the files and then goes through these within a nested loop. 

There all files are loaded with rtracklayer and a subst is made using GenomicRanges seqnames() function to only use the ranges listed as "chr1". These are then exported again using rtracklayer.

```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")
for (mark in marks) {
  print(mark)

  path <- "ENCODE_peaks/MARK/"
  currPath <- sub("MARK", mark, path)
  print(currPath)
  
  files <- list.files(path=currPath)
  for (file in files) {
    filePath <- paste(currPath, file, sep = "")
    peaks <- rtracklayer::import(filePath)
    chr1peaks <- peaks[seqnames(peaks) == "chr1"]
    
    chr1Path <- sub(".bigBed", "_chr1.bigBed", filePath)
    rtracklayer::export(chr1peaks, chr1Path)
  }
}
```
## Merging files within diseased and healthy groups

Here I am using the reduce function which will simply combine peaks if they overlap. An alternative would be disjoin.

Since I merge peaks the metadata is lost, but it is not valid anymore anyways. However rtracklayer expects a score value to be able to export to bigBed. So I manually assing each range a score of 1000.

```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")
h1 <- c("F27", "F43", "F27", "F27")
h2 <- c("F43", "M37", "M37", "M37")

for (i in 1:4) {
  print(marks[i])

  path <- "ENCODE_peaks/MARK/"
  currPath <- sub("MARK", marks[i], path)
  print(currPath)
  
  diseased1 <- paste(currPath, "MM.1S_", marks[i], "_GRCh38.bigBed", sep = "")
  diseased2 <- paste(currPath, "H929_", marks[i], "_GRCh38.bigBed", sep = "")
  print(diseased1)
  print(diseased2)
  peaksD1 <- rtracklayer::import(diseased1)
  peaksD2 <- rtracklayer::import(diseased2)
  
  # Merge peaks
  bothPeaks <- c(peaksD1, peaksD2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Diseased_", marks[i], "_GRCh38.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
  
  healthy1 <- paste(currPath, h1[i], "_", marks[i], "_GRCh38.bigBed", sep = "")
  healthy2 <- paste(currPath, h2[i], "_", marks[i], "_GRCh38.bigBed", sep = "")
  print(healthy1)
  print(healthy2)
  
  peaksH1 <- rtracklayer::import(healthy1)
  peaksH2 <- rtracklayer::import(healthy2)
  
  # Merge peaks
  bothPeaks <- c(peaksH1, peaksH2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Healthy_", marks[i], "_GRCh38.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
}

```
Due to lazyness the same for chr1
```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")
h1 <- c("F27", "F43", "F27", "F27")
h2 <- c("F43", "M37", "M37", "M37")

for (i in 1:4) {
  print(marks[i])

  path <- "ENCODE_peaks/MARK/"
  currPath <- sub("MARK", marks[i], path)
  print(currPath)
  
  diseased1 <- paste(currPath, "MM.1S_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  diseased2 <- paste(currPath, "H929_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  print(diseased1)
  print(diseased2)
  peaksD1 <- rtracklayer::import(diseased1)
  peaksD2 <- rtracklayer::import(diseased2)
  
  # Merge peaks
  bothPeaks <- c(peaksD1, peaksD2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Diseased_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
  
  healthy1 <- paste(currPath, h1[i], "_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  healthy2 <- paste(currPath, h2[i], "_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  print(healthy1)
  print(healthy2)
  
  peaksH1 <- rtracklayer::import(healthy1)
  peaksH2 <- rtracklayer::import(healthy2)
  
  # Merge peaks
  bothPeaks <- c(peaksH1, peaksH2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Healthy_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
}

```
## Finding poised sites

In order to find poised sites we look for overlaps in the peaks data between H3K4me3 and H3K27me3. OverlapsAny is not symetrical so I will try both ways. We will do this for heathly and diseased data separately.

Just like in the excercise we will use the subset of H3K4Me3, which is less generous.

Interestingly the poised sites seem to be more common in the healthy B-cells. Perhaps they have been lost. Usually this loss towards transcription is a transient response, but this could be different in cancer.

```{r}
disChr1H3K4Me3 <- rtracklayer::import("ENCODE_peaks/H3K4Me3/Diseased_H3K4Me3_GRCh38.bigBed")
disChr1H3K27Me3 <- rtracklayer::import("ENCODE_peaks/H3K27Me3/Diseased_H3K27Me3_GRCh38.bigBed")

test <- overlapsAny(disChr1H3K27Me3, disChr1H3K4Me3)
table(test)
test <- overlapsAny(disChr1H3K4Me3, disChr1H3K27Me3)
table(test)

rtracklayer::export(disChr1H3K4Me3[test], "ENCODE_peaks/Bivalent/Diseased_Bivalent_GRCh38.bed")
```

First lets see how many healthy bivalent sites overlap the ones in the diseased case.
We can see that a bit more than 38% of poised sites are still present in the diseased case
```{r}
bivDiseased <- rtracklayer::import("ENCODE_peaks/Bivalent/Diseased_Bivalent_GRCh38.bed")
bivHealthy <- rtracklayer::import("ENCODE_peaks/Bivalent/Healthy_Bivalent_GRCh38.bed")

bivalentOverlaps <- overlapsAny(bivHealthy, bivDiseased)
bivalentOverlaps <- bivHealthy[bivalentOverlaps]

length(bivalentOverlaps)
length(bivalentOverlaps) / length(bivHealthy) * 100
```



## Peak-wise comparisons

First lets just try an upset plot, Nevermind

```{r}
disChr1H3K4Me3 <- rtracklayer::import("ENCODE_peaks/H3K4Me3/Diseased_H3K4Me3_GRCh38_chr1.bigBed")
disChr1H3K9Me3 <- rtracklayer::import("ENCODE_peaks/H3K9Me3/Diseased_H3K9Me3_GRCh38_chr1.bigBed")
disChr1H3K27Ac <- rtracklayer::import("ENCODE_peaks/H3K27Ac/Diseased_H3K27Ac_GRCh38_chr1.bigBed")
disChr1H3K27Me3 <- rtracklayer::import("ENCODE_peaks/H3K27Me3/Diseased_H3K27Me3_GRCh38_chr1.bigBed")

helChr1H3K4Me3 <- rtracklayer::import("ENCODE_peaks/H3K4Me3/Healthy_H3K4Me3_GRCh38_chr1.bigBed")
helChr1H3K9Me3 <- rtracklayer::import("ENCODE_peaks/H3K9Me3/Healthy_H3K9Me3_GRCh38_chr1.bigBed")
helChr1H3K27Ac <- rtracklayer::import("ENCODE_peaks/H3K27Ac/Healthy_H3K27Ac_GRCh38_chr1.bigBed")
helChr1H3K27Me3 <- rtracklayer::import("ENCODE_peaks/H3K27Me3/Healthy_H3K27Me3_GRCh38_chr1.bigBed")

peaks <- list(disChr1H3K4Me3, disChr1H3K9Me3, disChr1H3K27Ac, disChr1H3K27Me3)
names(peaks) <- c("1", "2", "3", "4")
regionUpset(peaks)
```

We first want a table to see how many peaks overlap between diseased and healthy B-cells for a mark. Doing overlaps any and then printing the table of true and false values gives us thid data.

We have about a 50:50 split of peaks H3K4Me3, meaning that we can find around 50% of peaks only in the diseased models and not in the healthy ones.

We now want to export the disease-exclusive peaks, but we can not use the overlaps values to subset the diseased peaks directly. Instead we want to invert the *TRUE* and *FALSE* values, so that we get the peaks that do not overlap. In R this can be done conveniently with the *!* operator.

Then we export these peaks as unique peaks. Here I suddenly got problems as there were two columns with strand information. One in the format * and one in . format

```{r}
disChr1H3K4Me3 <- rtracklayer::import("ENCODE_peaks/H3K4Me3/Diseased_H3K4Me3_GRCh38_chr1.bigBed")
helChr1H3K4Me3 <- rtracklayer::import("ENCODE_peaks/H3K4Me3/Healthy_H3K4Me3_GRCh38_chr1.bigBed")

overlaps <- overlapsAny(disChr1H3K4Me3, helChr1H3K4Me3)
table(overlaps)

nonOverlaps <- !overlaps
table(nonOverlaps)

nonOverlapsPeaks <- disChr1H3K4Me3[nonOverlaps]
length(nonOverlapsPeaks)

rtracklayer::export(nonOverlapsPeaks, "ENCODE_peaks/H3K4Me3/Diseased_Exclusive_H3K4Me3_GRCh38_chr1.bed")
```
Automated version of the pipeline:
Results are interesting, for most marks there is also around a 50:50 split. But not for H3K27Me3. There we find a much higher proportion of exclusive peaks.

```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")

for (mark in marks) {
  diseasedPath <- "ENCODE_peaks/MARK/Diseased_MARK_GRCh38_chr1.bigBed"
  diseasedPath <- gsub("MARK", mark, diseasedPath)
  print(diseasedPath)
  diseasedPeaks <- rtracklayer::import(diseasedPath)
  
  healthyPath <- "ENCODE_peaks/MARK/Healthy_MARK_GRCh38_chr1.bigBed"
  healthyPath <- gsub("MARK", mark, healthyPath)
  print(healthyPath)
  healthyPeaks <- rtracklayer::import(healthyPath)
  
  overlaps <- overlapsAny(diseasedPeaks, healthyPeaks)
  print(table(overlaps))

  nonOverlaps <- !overlaps
  print(table(nonOverlaps))

  nonOverlapsPeaks <- diseasedPeaks[nonOverlaps]
  print(length(nonOverlapsPeaks))

  exportPath <- "ENCODE_peaks/MARK/Diseased_Exclusive_MARK_GRCh38_chr1.bed"
  exportPath <- gsub("MARK", mark, exportPath)
  rtracklayer::export(nonOverlapsPeaks, exportPath)
}
```

Doing it the other way for peaks exclusive to healthy B-cells:
```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")

for (mark in marks) {
  diseasedPath <- "ENCODE_peaks/MARK/Diseased_MARK_GRCh38_chr1.bigBed"
  diseasedPath <- gsub("MARK", mark, diseasedPath)
  print(diseasedPath)
  diseasedPeaks <- rtracklayer::import(diseasedPath)
  
  healthyPath <- "ENCODE_peaks/MARK/Healthy_MARK_GRCh38_chr1.bigBed"
  healthyPath <- gsub("MARK", mark, healthyPath)
  print(healthyPath)
  healthyPeaks <- rtracklayer::import(healthyPath)
  
  overlaps <- overlapsAny(healthyPeaks, diseasedPeaks)
  print(table(overlaps))

  nonOverlaps <- !overlaps
  print(table(nonOverlaps))

  nonOverlapsPeaks <- healthyPeaks[nonOverlaps]
  print(length(nonOverlapsPeaks))

  exportPath <- "ENCODE_peaks/MARK/Healthy_Exclusive_MARK_GRCh38_chr1.bed"
  exportPath <- gsub("MARK", mark, exportPath)
  rtracklayer::export(nonOverlapsPeaks, exportPath)
}
```
## Basic Statistics

Overall amount of peaks. I noticed that we have more peaks overall in healthy cases, so lets calculate how many more. The following statistics are based on the genome-wide peaks.

```{r}
# H3K4Me3, H3K9Me3, H3K27Ac, H3K27Me3
diseasedAmount <- c(47173, 13749, 36917, 56005)
healthyAmount <- c(32884, 155871, 50871, 98133)
totalDif <- c()

for (i in 1:4) {
  totalDif <- c(totalDif, healthyAmount[i] - diseasedAmount[i])
}
sum(totalDif) / 4
totalAmount <- sum(diseasedAmount) + sum(healthyAmount)
sum(totalDif) / totalAmount
```


From the information generated before we can do some basic statistics
data structure: 
- unique diseased peaks
- unique healthy peaks
- peaks shared
```{r}
#par(mfrow=c(2,2))
H3K4Me3 <- c(24631, 9792, 47173+32884-24631-9792)
pie(H3K4Me3, main = "H3K4Me3", c("exclusive diseased", "exclusive healthy", "shared peaks"), col=c("springgreen4", "springgreen3", "springgreen2"))

H3K9Me3 <- c(6796, 149416, 13749+155871-6796-149416)
pie(H3K9Me3, main = "H3K9Me3", c("exclusive diseased", "exclusive healthy", "shared peaks"), col=c("orangered4", "orangered3", "orangered2"))

H3K27Ac <- c(19656, 31592, 36917+50871-19656-31592)
pie(H3K27Ac, main = "H3K27Ac", c("exclusive diseased", "exclusive healthy", "shared peaks"), col=c("springgreen4", "springgreen3", "springgreen2"))

H3K27Me3 <- c(39445, 76377, 56005+98133-39445-76377)
pie(H3K27Me3, main = "H3K27Me3", c("exclusive diseased", "exclusive healthy", "shared peaks"), col=c("orange4", "orange3", "orange2"))

Bivalent <- c(3104, 3891, 4773+6342-3104-3891)
pie(Bivalent, main = "Bivalent", c("exclusive diseased", "exclusive healthy", "shared peaks"), col=c("orange4", "orange3", "orange2"))
```

