---
title: "3_Peak Comparison"
author: "Laurin Schaller"
date: "2023-06-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(rtracklayer)
  library(GenomicRanges)
})
```


## Subsetting to chromosome 1

First we want to subset all the bigBed-files we downloaded from ENCODE to chromosome 1 for reduced sizes.
The following function uses a pythonic approach to looping all the histone marks which have  the same folder names. Within these folders it gets all the files and then goes through these within a nested loop. 

There all files are loaded with rtracklayer and a subst is made using GenomicRanges seqnames() function to only use the ranges listed as "chr1". These are then exported again using rtracklayer.

```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")
for (mark in marks) {
  print(mark)

  path <- "ENCODE_peaks/MARK/"
  currPath <- sub("MARK", mark, path)
  print(currPath)
  
  files <- list.files(path=currPath)
  for (file in files) {
    filePath <- paste(currPath, file, sep = "")
    peaks <- rtracklayer::import(filePath)
    chr1peaks <- peaks[seqnames(peaks) == "chr1"]
    
    chr1Path <- sub(".bigBed", "_chr1.bigBed", filePath)
    rtracklayer::export(chr1peaks, chr1Path)
  }
}
```
## Merging files within diseased and healthy groups

Here I am using the reduce function which will simply combine peaks if they overlap. An alternative would be disjoin.

Since I merge peaks the metadata is lost, but it is not valid anymore anyways. However rtracklayer expects a score value to be able to export to bigBed. So I manually assing each range a score of 1000.

```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")
h1 <- c("F27", "F43", "F27", "F27")
h2 <- c("F43", "M37", "M37", "M37")

for (i in 1:4) {
  print(marks[i])

  path <- "ENCODE_peaks/MARK/"
  currPath <- sub("MARK", marks[i], path)
  print(currPath)
  
  diseased1 <- paste(currPath, "MM.1S_", marks[i], "_GRCh38.bigBed", sep = "")
  diseased2 <- paste(currPath, "H929_", marks[i], "_GRCh38.bigBed", sep = "")
  print(diseased1)
  print(diseased2)
  peaksD1 <- rtracklayer::import(diseased1)
  peaksD2 <- rtracklayer::import(diseased2)
  
  # Merge peaks
  bothPeaks <- c(peaksD1, peaksD2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Diseased_", marks[i], "_GRCh38.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
  
  healthy1 <- paste(currPath, h1[i], "_", marks[i], "_GRCh38.bigBed", sep = "")
  healthy2 <- paste(currPath, h2[i], "_", marks[i], "_GRCh38.bigBed", sep = "")
  print(healthy1)
  print(healthy2)
  
  peaksH1 <- rtracklayer::import(healthy1)
  peaksH2 <- rtracklayer::import(healthy2)
  
  # Merge peaks
  bothPeaks <- c(peaksH1, peaksH2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Healthy_", marks[i], "_GRCh38.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
}

```
Due to lazyness the same for chr1
```{r}
marks <- c("H3K4Me3", "H3K9Me3", "H3K27Ac", "H3K27Me3")
h1 <- c("F27", "F43", "F27", "F27")
h2 <- c("F43", "M37", "M37", "M37")

for (i in 1:4) {
  print(marks[i])

  path <- "ENCODE_peaks/MARK/"
  currPath <- sub("MARK", marks[i], path)
  print(currPath)
  
  diseased1 <- paste(currPath, "MM.1S_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  diseased2 <- paste(currPath, "H929_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  print(diseased1)
  print(diseased2)
  peaksD1 <- rtracklayer::import(diseased1)
  peaksD2 <- rtracklayer::import(diseased2)
  
  # Merge peaks
  bothPeaks <- c(peaksD1, peaksD2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Diseased_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
  
  healthy1 <- paste(currPath, h1[i], "_", marks[i], "_GRCh38.bigBed_chr1", sep = "")
  healthy2 <- paste(currPath, h2[i], "_", marks[i], "_GRCh38.bigBed_chr1", sep = "")
  print(healthy1)
  print(healthy2)
  
  peaksH1 <- rtracklayer::import(healthy1)
  peaksH2 <- rtracklayer::import(healthy2)
  
  # Merge peaks
  bothPeaks <- c(peaksH1, peaksH2)
  mergedPeaks <- reduce(bothPeaks)
  mergedPeaks$score <- 1000
  
  exportPath <- paste(currPath, "Healthy_", marks[i], "_GRCh38_chr1.bigBed", sep = "")
  rtracklayer::export(mergedPeaks, exportPath)
}

```

